A weighted sampling algorithm to designing RNA sequences with target secondary structure and nucleotide distribution

### 2013-01-08 ISMB/main_ISMB.tex
notation

\newcommand{\ourprog}{\texttt{Incarnation}\xspace}

@@

\begin{document}
\maketitle
%\section{}
%\subsection{}
\begin{abstract}
\noindent
\textbf{Motivations:} The design of RNA sequences folding into predefined secondary structures is a milestone for many synthetic biology and gene therapy studies. Most of the current software use similar local search strategies (i.e. a random seed is progressively adapted to acquire the desired folding propertues) and more importantly do not allow the user to control explicitly the nucleotide distribution such as the \gc~content in their sequences. However, the latter is an important criteria for large-scale applications as it could presumably be used to design sequences with better transcription rates and/or structural plasticity.\\
\noindent
\textbf{Results:}
In this paper, we introduce \ourprog, a novel algorithm to design RNA sequences folding into target secondary structures with a predefined nucleotide distribution.  \ourprog uses a global sampling approach and weighted sampling techniques. We show that our approach is fast (i.e. running time comparable or better than local search methods), seed-less (we remove the bias of the seed in local search heuristics), and successfully generates high-quality sequences (i.e. thermodynamically stable) for any \gc~content. To complete this study, we develop an hybrid method combining our global sampling approach with local search strategies. Remarkably, our glocal methodology overcomes both local and global approaches.\\
\noindent
\textbf{Key words:} RNA, secondary structure, design, weighted sampling, GC-content.
\end{abstract}

>>>>

\newcommand{\ourprog}{\texttt{IncaRNAtion}\xspace}

@@

\begin{document}
\maketitle
%\section{}
%\subsection{}
\begin{abstract}
\noindent
\textbf{Motivations:} The design of RNA sequences folding into predefined secondary structures is a milestone for many synthetic biology and gene therapy studies. Most of the current software use similar local search strategies (i.e. a random seed is progressively adapted to acquire the desired folding propertues) and more importantly do not allow the user to control explicitly the nucleotide distribution such as the \GCContent in their sequences. However, the latter is an important criteria for large-scale applications as it could presumably be used to design sequences with better transcription rates and/or structural plasticity.\\
\noindent
\textbf{Results:}
In this paper, we introduce \ourprog, a novel algorithm to design RNA sequences folding into target secondary structures with a predefined nucleotide distribution.  \ourprog uses a global sampling approach and weighted sampling techniques. We show that our approach is fast (i.e. running time comparable or better than local search methods), seed-less (we remove the bias of the seed in local search heuristics), and successfully generates high-quality sequences (i.e. thermodynamically stable) for any \GCContent. To complete this study, we develop an hybrid method combining our global sampling approach with local search strategies. Remarkably, our glocal methodology overcomes both local and global approaches.\\
\noindent
\textbf{Key words:} RNA, secondary structure, design, weighted sampling, GC-content.
\end{abstract}

### 2013-01-08 ISMB/results_ISMB.tex
typos;
notation

\subsection{Implementation}
\label{sec:implementation}
Our software, \texttt{Incarnation}, was implemented in {\tt Python~2.7}. We used
\RNAinverse from the \textit{Vienna Package 2.0}~\cite{Hofacker:1994}.
All time benchmarks were run on a one AMD Opteron(tm) Processor 6278  at 2.4 GHz with cache of 512 KB.
 We used as parameter $\beta=15$, the penalty for an invalid base pair.

We present the average times spent by \texttt{Incarnation} and \RNAinverse to generate one sequence
with the required \GCContent in Figure~\ref{fig:time}. As expected, the time grows linearly
in function of the length of the structures for \texttt{Incarnation}.  A highly interesting feature
is the huge decrease in time of computation when there is $15\%$ free 
nucleotides inside the loops of the secondary structure.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/time_rnastrand_clustered_rnainverse_100samples_fix}
	\caption{Average time in seconds to generate one sequence. We explicitly show 
	the time spent by \texttt{Incarnation} (full line) and \texttt{RNAinverse} (dashed) for various \GCContent. The first plot is in function
	of the length of the structures. The second is in function of the stack
	density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	the proportion of free nucleotides (i.e. not base paired) inside loops.}
	\label{fig:time}	
\end{figure}

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \texttt{Incarnation} in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \texttt{Incarnation} sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is not a good predictor for the quality of the results. Instead,
a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
 The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last column of Figure~\ref{fig:mfe_struct_solved}.
Even with a \GCContent of $50\%$ and more, when at least
one solutions was found for almost all structures, most of the samples 
do not fold into the target. A few structures remained unsolved under
all \GCContent.

\begin{figure}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solve_nornainverse.png}
	\caption{Quality of \texttt{Incarnation} results. The first row shows the percentage
	of sampled sequences folding into the target. The second shows the 	
	proportion	of structures for which at least one correct sequence was 
	sampled.}
	\label{fig:mfe_struct_solved_noinverse}	
\end{figure}

@@

To evaluate the global quality of \texttt{Incarnation} sequences, we show
in Figure~\ref{fig:ss_sens} the ratio of well predicted base pairs in the
MFE structure of our sampled sequences. We can observe that, in all cases,
the hardest sequence to design are those with an extremely low \GCContent. As anticipated, those are the sequences with the weakest bonds.
We notice that the most accurate sequences yield a \GCContent
of $70\pm 10\%$.

@@

A desirable feature in sequence design, is to produce samples with a high
diversity. We show in Figure~\ref{fig:nb_sols_entropy} the number of correct
solutions as their average entropy and base pair entropy, since 
\texttt{Incarnation}  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \GCContent and the percentage of free nucleotides. Our method is able to generate

@@

\subsection{Global sampling vs Local search vs Glocal approach}
Here, we are interested in estimating the impact of the design methodology on the performances. More precisely, we aim to what are the benefits of a global sampling approach (\ourprog), versus a glocal procedure (\ourprog + \RNAinverse) or the classical local search methodology (\RNAinverse). Thus, we compare our method to a pure local search procedure. In particular, here we compare to \RNAinverse as it is the program we used to do the local optimization (see Section \ref{subsec:glocal_method}). Our results (See Fig. \ref{fig:rnainverse} in Supplementary data) show that \ourprog overcome other strategies. In particular, the base pair entropy as well as diversity of sequences are favorable to \ourprog. This advantage is even more striking for low \gc~content regimes.

>>>>

\subsection{Implementation}
\label{sec:implementation}
Our software, \ourprog, was implemented in {\tt Python~2.7}. We used
\RNAinverse from the \textit{Vienna Package 2.0}~\cite{Hofacker:1994}.
All time benchmarks were run on a one AMD Opteron(tm) Processor 6278  at 2.4 GHz with cache of 512 KB.
 We used as parameter $\beta=15$, the penalty for an invalid base pair.

We present the average times spent by \ourprog and \RNAinverse to generate one sequence
with the required \GCContent in Figure~\ref{fig:time}. As expected, the time grows linearly
in function of the length of the structures for \ourprog.  A highly interesting feature
is the huge decrease in time of computation when there is $15\%$ free 
nucleotides inside the loops of the secondary structure.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/time_rnastrand_clustered_rnainverse_100samples_fix}
	\caption{Average time in seconds to generate one sequence. We explicitly show 
	the time spent by \ourprog (full line) and \texttt{RNAinverse} (dashed) for various \GCContent. The first plot is in function
	of the length of the structures. The second is in function of the stack
	density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	the proportion of free nucleotides (i.e. not base paired) inside loops.}
	\label{fig:time}	
\end{figure}

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \ourprog in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \ourprog sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is not a good predictor for the quality of the results. Instead,
a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
 The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last column of Figure~\ref{fig:mfe_struct_solved}.
Even with a \GCContent of $50\%$ and more, when at least
one solutions was found for almost all structures, most of the samples 
do not fold into the target. A few structures remained unsolved under
all \GCContent.

\begin{figure}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solve_nornainverse.png}
	\caption{Quality of \ourprog results. The first row shows the percentage
	of sampled sequences folding into the target. The second shows the 	
	proportion	of structures for which at least one correct sequence was 
	sampled.}
	\label{fig:mfe_struct_solved_noinverse}	
\end{figure}

@@

To evaluate the global quality of \ourprog sequences, we show
in Figure~\ref{fig:ss_sens} the ratio of well predicted base pairs in the
MFE structure of our sampled sequences. We can observe that, in all cases,
the hardest sequence to design are those with an extremely low \GCContent. As anticipated, those are the sequences with the weakest bonds.
We notice that the most accurate sequences yield a \GCContent
of $70\pm 10\%$.

@@

A desirable feature in sequence design, is to produce samples with a high
diversity. We show in Figure~\ref{fig:nb_sols_entropy} the number of correct
solutions as their average entropy and base pair entropy, since 
\ourprog  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \GCContent and the percentage of free nucleotides. Our method is able to generate

@@

\subsection{Global sampling vs Local search vs Glocal approach}
Here, we are interested in estimating the impact of the design methodology on the performances. More precisely, we aim to what are the benefits of a global sampling approach (\ourprog), versus a glocal procedure (\ourprog + \RNAinverse) or the classical local search methodology (\RNAinverse). Thus, we compare our method to a pure local search procedure. In particular, here we compare to \RNAinverse as it is the program we used to do the local optimization (see Section \ref{subsec:glocal_method}). Our results (See Fig. \ref{fig:rnainverse} in Supplementary data) show that \ourprog overcome other strategies. In particular, the base pair entropy as well as diversity of sequences are favorable to \ourprog. This advantage is even more striking for low \GCContent regimes.
